---
title: "Control-Flow Constructs"
author: "Introduction to R"
output:
  html_document:
    theme: spacelab
---

## Question 1

Write an `if-else` construct that prints the statement "a < A" if `"a" < "A"` is `TRUE` and "a >= A" otherwise.
```{r}
# FILL ME IN
if ("a"<"A")
{
  print("a < A")
}else
{
  print("a >= A")
}

```

## Question 2

Repeat Q1, but using the `ifelse()` function.
```{r}
# FILL ME IN
ifelse("a"<"A","a < A","a >= A")
```

## Question 3

Write an `if` construct that, if there are matching elements of `u` and `v` (i.e., if `u[i] = v[i]` for any index i), prints the number of matching elements, and otherwise prints the string "There are no matching elements."
```{r}
# FILL ME IN
u <- 1:50
v <- c(99,105,35,23,77,91,6,57,41)
equal <- table(match(u,v))
if (length(equal>0)){
  print(length(equal))
}else{
  print("There are no matching elements")
}

```

## Question 4

Confirm the short-circuiting behavior of an `if-else` construct by (a) setting the variable `t` to the value 4, and (b) writing an `if-else` construct that first checks if `t` is greater than or equal to 4 and if so prints "t >= 4", then checks to see if `t` is greater than zero and if so prints "t > 0", and otherwise prints "t <= 0".
```{r}
# FILL ME IN
t<-4
if(t>=4){
  print("t >= 4")
}else if (t > 0){
  print("t > 0")
}else {
  print("t <= 0")
}
```

