---
title: "All About Lists"
author: "Introduction to R"
output:
  html_document:
    theme: spacelab
---

## Question 1

Create an empty list $x$. Then define its *second* entry as the vector 2:4. Then print the list. Note what value does the first entry default to.
```{r}
# FILL ME IN
x<-list()
x
x[[2]]<-2:4
x
```

## Question 2

Use the `names()` function to rename the list entries to `x` and `y`. Print $x$ to ensure your changes took hold.
```{r}
# FILL ME IN
names(x)<-c("x","y")
x
```

## Question 3

Change the name of the first entry of the list $x$ to `a`. 
```{r}
# FILL ME IN
names(x)[1]<-c("a")
x
```

## Question 4

Create a data frame `df` that has columns `x` and `y` and has three rows. Use the `nrow()`, `ncol()`, and `dim()` functions to display the number of rows, the number of columns, and the dimensions of `df`. Let the first column contain numbers, and the second column contain logical values.
```{r}
# FILL ME IN
df<-data.frame(x=1:3,y=c(TRUE,TRUE,FALSE))
df
nrow(df)
ncol(df)
dim(df)
```

## Question 5

Add columns to `df` using the dollar sign operator, using the double bracket notation with number, and using the double bracket notation with character name.
```{r}
# FILL ME IN
df$z<-c(8,9,5)
df[[4]]<-c("a","b","c")
df[["a"]]<-11:13
df

```

## Question 6

Use `row.names()` to change the names of the rows of `df` to "1st", "2nd", and "3rd". (It works the same way `names()` does.)
```{r}
# FILL ME IN
row.names(df)<-c("1st","2nd","3rd")
df[1:5]
```

## Question 7

Display the contents of the first row of `df` using the row number and then using the row name. Note that you access the elements of a two-dimensional object using `[row number/name,column number/name]`.
```{r}
# FILL ME IN
df[1,]
df["1st",]
```

## Question 8

Initialize a 2 x 2 matrix where all the matrix elements are 1. Display the matrix.
```{r}
# FILL ME IN
mat<-matrix(1,nrow=2,ncol=2)
mat
```

