---
title: "Loops"
author: "Introduction to R"
output:
  html_document:
    theme: spacelab
---

## Question 1

Write a `for()` loop to add, elementwise, the vectors `x` and `y` defined below. Place each sum into the vector `x.plus.y` (which you can initialize using, e.g., `x.plus.y <- rep(NA,4)`, where `NA` means "not available" or missing), then after the `for()` loop, display `x.plus.y`.
```{r}
x <- c(1,2,3,4)
y <- c(-2,2,-3,3)
# FILL ME IN
x.plus.y <- rep(NA,4)
for(ii in seq_along(x.plus.y)){
  x.plus.y[ii] <- x[ii]+y[ii]
}
x.plus.y
```

## Question 2

Write a `for()` loop to sum the natural logarithms of all elements of the vector `z`, defined below, that are positive. (Use `if` to check for positivity.) Display the result. Show that you can perform the same task without a `for()` loop, by utilizing logical-based vector subsetting and one call to `sum()`.
```{r}
z <- c(-5,1,2,-4,3,4,-3,6)
sum.z.positive <- 0
# FILL ME IN
for (ii in 1:length(z)){
  if (z[ii] >0) {
    sum.z.positive <- sum.z.positive + log(z[ii])
  }
}
sum.z.positive
sum(log(z[z>0]))
```

## Question 3

Write a `while()` loop that computes the sum of the first 100 positive integers. (Set the variable `s` equal to zero, then increment its value with each loop.) Then display the result. (It should be 5050.)
```{r}
# FILL ME IN
condition <-TRUE
sumofhundred <- 0
num <- 1
while(condition){
  if (num <=100){
    sumofhundred = sumofhundred + num
  }else {
    condition <- FALSE
  }
  num <- num+1
}
sumofhundred
```

## Question 4

Repeat Q3, but use a `for()` loop instead.
```{r}
# FILL ME IN
sumofhundred <- 0
for (num in 1:100){
  sumofhundred = sumofhundred + num
}
sumofhundred
```

## Question 5

Write a `while()` loop that samples one value from a standard normal, then breaks when the value is greater than 4. (The code would be `if ( rnorm(1) > 4 ) break`.) Also include an incrementing variable that increments by one with each loop, and display its value when the loop is broken. (In other words: display how many loops occurred before a sampled value of > 4 was observed.)
```{r}
# FILL ME IN
count <- 1
while(count){
  if (rnorm(1) > 4) break
  count <- count +1
}
count
  
```

