---
title: "All About Vectors"
author: "Introduction to R"
output:
  html_document:
    theme: spacelab
---

# Preliminaries

After each question, you will see the following:
```{r}
# FILL ME IN
```
This, in `R Markdown` parlance, is a "code chunk." To answer the question, replace this line with your answer. Note that anything following a "#" symbol is a comment (or is code that is "commented out"). Also note that you do not need to remove the question or make other edits. Just fill in the code chunks.

To run the chunk to see if it works, simply put your cursor *inside* the chunk and, e.g., select "Run Current Chunk" from the "Run" pulldown tab. Alternately, you can click on the green arrow at the upper right-most part of the chunk, or use "\<cntl\>-\<return\>" as a keyboard shortcut.

Here is an example (that only makes sense if you are looking at the `.Rmd` file):

## Question 0

Print "Hello, world."
```{r}
print("Hello, world.")
```

When you have finished answering the questions, you can click on the "Knit" button to see a "compiled" version of your answers. Note that `R Markdown` may prompt you to install packages to get the knitting to work; do install these.

---

## Question 1

Initialize a vector $y$ with one logical value, one numeric value, and one character value, and determine the type of $y$.

```{r}
# FILL ME IN
y<- c(TRUE, 8, 'A')
typeof(y)
```

## Question 2

Sort the vector $y$ into ascending order. Note how the data are ordered.
```{r}
# FILL ME IN
sort(y)
```

## Question 3

Initialize a vector $y$ of integers, with first value 4 and last value -4, stepping down by 1. Do this *two* different ways. After each initialization, print the vector.
```{r}
# FILL ME IN
y<- seq(4, -4, by=-1)
y
y<-4:-4
y

```

## Question 4

What variable type is 1?
Divide 1 by 2. Note to yourself whether you get zero or 0.5.
```{r}
# FILL ME IN
x<-1
typeof(x)
d<- (1/2)
d
```

## Question 5

Take the vector $x$ defined below and display the elements that are less than $-1$ or greater than $1$. Do this using the logical or symbol, and again via the use of the `abs()` function (for absolute value).
```{r}
set.seed(199)
x <- rnorm(20)
x[x< -1 | x>1]
x[abs(x)>1]

# FILL ME IN
```

## Question 6

What proportion of values in the vector $x$ are less than 0.5? Use `sum()` and `length()` in your answer.
```{r}
# FILL ME IN
sum(x<0.5)/length(x)
```

## Question 7

Sort all the values of $x$ in *decreasing* order. Do this two different ways.
To see the help page for `sort()` (which might help you), type `?sort` at
the Console prompt.
```{r}
# FILL ME IN
sort(x, decreasing = TRUE)
rev(sort(x))
```

## Question 8

Replace all positive values in the vector $x$ with zero, using `which()`.
```{r}
# FILL ME IN
x
w<-which(x>0)
x[w]<-0
x
```

