---
title: "Handy Vector Functions"
author: "Introduction to R"
output:
  html_document:
    theme: spacelab
---

Here we define some vectors:
```{r}
set.seed(1201)
u <- sample(100,100,replace=TRUE)
v <- sample(100,100,replace=TRUE)
l <- list("x"=sample(1:10,5),"y"=sample(11:20,5))
df <- data.frame("x"=sample(1:10,5),"y"=sample(11:20,5))
x <- c(1,2,3,4)
y <- c(-2,2,-3,3)
z <- c(-5,1,2,-4,3,4,-3,6)
```

## Question 1

Display the list `l` as a numerical vector, with names associated with each element.
```{r}
# FILL ME IN
unlist(l)
```

## Question 2

Display the list `l` as a numerical vector, while stripping away the names seen in Q1.
```{r}
# FILL ME IN
vect<-as.vector(unlist(l))
vect
```

## Question 3

Repeat Q2, but display the vector in *descending* order.
```{r}
# FILL ME IN
sort(vect,decreasing=TRUE)


```

## Question 4

Here are the contents of the data frame `df`:
```{r}
df
```
Reorder the rows so that the entries of the `x` column are in numerical order and the association between the i<sup>th</sup> entry of `x` and the i<sup>th</sup> entry of `y` is not lost. Display the result.
```{r}
# FILL ME IN
a<-df[["x"]]
b<-df[["y"]]
o<-order(a)
a[o]
b[o]

```

## Question 5

Display the proportion of the total number of unique values in `u` to the number of values in `u`.
```{r}
# FILL ME IN
length(unique(u))/length(u)
```

## Question 6

Display a table that shows how often each value of `v` appears.
```{r}
# FILL ME IN
table(v)
```

## Question 7

How many unique values do `u` and `v` have in common?
```{r}
# FILL ME IN
length(intersect(u,v))
```

## Question 8

Write down an expression that returns `TRUE` if the union of `u` and `v` has 100 elements and `FALSE` otherwise.
```{r}
# FILL ME IN
length(union(u,v)) == 100
```

## Question 9

Display the (sorted!) values of `u` that do not appear in `v`.
```{r}
# FILL ME IN
setdiff(sort(u),v)
```

## Question 10

Display a table showing how many values that are in `v` but not in `u` fall into the bins [1,50] and [51,100].
```{r}
# FILL ME IN
s<-setdiff(v,u)
table(findInterval(s,c(51)))

```
